---
title: "Data Exploration and Preprocessing"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# ProblemA.R
library(ggplot2)
library(tidyverse)
library(lubridate)

ratings_from_csv <- function(file_path) {
    ratings <- read.csv(file_path, header=FALSE, sep='\t')
    names(ratings) <- c('user_id', 'movie_id', 'rating', 'timestamp')
    return(ratings)
}

date_to_sec <- function(date_str) {
  as.integer(as.POSIXct(as.Date(date_str, format='%d-%b-%Y')))
}

# Create data frames for ratings
ratings_train <- ratings_from_csv('/Users/Stephanie/Desktop/OMSI/omsi/ml-100k/ua.base')
ratings_test <- ratings_from_csv('/Users/Stephanie/Desktop/OMSI/omsi/ml-100k/ua.test')
ratings <- rbind(ratings_train, ratings_test)

# Create data frame for movie data
movie_info <- read.csv('/Users/Stephanie/Desktop/OMSI/omsi/ml-100k/u.item', header=FALSE, sep='|')
names(movie_info) <- c('movie_id', 'movie_title', 'release_date',
                       'video_release_date', 'imdb_url', 'unknown', 'action',
                       'adventure', 'animation', 'childrens', 'comedy', 'crime',
                       'documentary', 'drama', 'fantasy', 'filmnoir', 'horror',
                       'musical', 'mystery', 'romance', 'scifi', 'thriller',
                       'war', 'western')

# Create data frame for user data
user_info <- read.csv('/Users/Stephanie/Desktop/OMSI/omsi/ml-100k/u.user', header=FALSE, sep='|')
names(user_info) <- c('user_id', 'age', 'gender', 'occupation', 'zip_code')

# Merge movie user data into rating data
ratings <- merge(ratings, movie_info, by.x='movie_id', by.y='movie_id')
ratings <- merge(ratings, user_info, by.x='user_id', by.y='user_id')
head(ratings)
```

## Preprocessing
### Missing Values
```{r}
NA_ratings <- sapply(ratings, function(x) switch( class(x), factor = sum(x==""), sum( is.na(x) ) ))
NA_ratings_df <- as.data.frame( t(NA_ratings) )
NA_ratings_df[which(NA_ratings_df != 0)]
```
From above, we can see that all values of video_release_date are NA. This suggests that we can remove the feature from our dataset without losing any information.

We can also see that there are 9 observations with missing values. We can take a closer look to at the overvations that have NA value for release date.

```{r}
# Convert movie release date to seconds from UNIX epoch
ratings$release_date <- apply(ratings[c('release_date')], 1, date_to_sec)
ratings %>% filter(release_date %>% is.na())
```
From above, we can see that the observations with missing values for release_date are also missing information about movie_title and genre. This missing information will make these observations not very useful for rating predictions so we will drop them.

```{r}
# remove overvations with missing values for release_date
missing_dates <- ratings$release_date %>% is.na() %>% which()
ratings <- ratings[-missing_dates,]
```

The final variable with missing values is imdb_url. We will be removing this feature because it does not give useful information related to movie rating. 

### Feature Removal
We decided to remove the three following features: movie_title, video_release_date, and imdb_url. As discussed above, we will remove video_release_date because it have all missing values and imdb_url because it has missing values and is noninformative in regards to movie rating. Lastly, we will remove movie_title because it is redundant information since we already have movie_id.

```{r}
# Remove noninfomative predictors
drops <- c('movie_title','video_release_date','imdb_url')
ratings <- ratings[ , !names(ratings) %in% drops]
```

### Feature Preprocessing
Preprocess timestamp and release_date to be consistent
```{r}
# convert timestamp and release date to class Date for fprocessing later
ratings$timestamp <- ratings$timestamp %>% as_datetime
ratings$release_date <- ratings$release_date %>% as_datetime
```

zip_code 
```{r}
# replace old zipcode column with two digits
ratings$zip_code <- substr(as.character(ratings$zip_code),1,2)
```

Convert categorical variables to factors
```{r}
# convert all categorical variables to factors
factor_cols <- c('user_id', 'movie_id', 'unknown', 'action',
                       'adventure', 'animation', 'childrens', 'comedy', 'crime',
                       'documentary', 'drama', 'fantasy', 'filmnoir', 'horror',
                       'musical', 'mystery', 'romance', 'scifi', 'thriller',
                       'war', 'western','zip_code')
ratings[,factor_cols] <- data.frame(apply(ratings[factor_cols], 2, as.factor))
```

### Feature Engineering
Time intervals
```{r}
ratings$release_year <- year(ratings$release_date)
ratings$release_month <- month(ratings$release_date)
ratings$timestamp_year <- year(ratings$timestamp)
ratings$timestamp_month <- month(ratings$timestamp)
ratings$time_difference <- as.period(ratings$timestamp - ratings$release_date) %>% day
```

Age intervals
```{r}
age <- ratings %>% pull(age)
ratings$age_group <- rep(0,nrow(ratings))
ratings$age_group <- findInterval(age,c(10,20,30,40,50,60,70,80))
ratings$age_group <- as.factor(ratings$age_group)
# 0, 1, 2, 3, 4, 5, 6, 7
levels(ratings$age_group) <- c("<10","10-20","20-30","30-40","40-50","50-60","60-70","70+")
```


## Summary Statistics ########################################################
Distibution of Feature Values
```{r}
summary(ratings)
```

Features Types and Values
```{r}
str(ratings)
```

### Gender
Number of ratings by gender
```{r, echo=FALSE}
summary(ratings$gender)
```

Rating Proportions by Gender
```{r, echo=FALSE}
prop.table(table(ratings$gender, ratings$rating))
```

Row wise rating Proportions
```{r, echo=FALSE}
prop.table(table(ratings$gender, ratings$rating),1)
```

### Age
Number of ratings by age
```{r, echo=FALSE}
summary(ratings$age)
```

Rating proportions by age group
```{r, echo=FALSE}
age <- ratings %>% pull(age)
age_group <- as.factor(findInterval(age,c(10,20,30,40,50,60,70,80)))
levels(age_group) <- c("<10","10-20","20-30","30-40","40-50","50-60","60-70","70+")
prop.table(table(age_group, ratings$rating))
```

Row wise rating proportions by age group
```{r, echo=FALSE}
prop.table(table(age_group, ratings$rating))
```

Table with both gender and age
```{r}
aggregate(rating ~ age_group + gender, data=ratings, FUN=sum)
```

```{r}
aggregate(rating ~ age_group + gender, data=ratings, FUN=length)
```

```{r}
aggregate(rating ~ age_group + gender, data=ratings, FUN=function(x) {length(x)/sum(x)})
```

```{r}
aggregate(rating ~ zip_code + age_group + gender, data=ratings, FUN=function(x) {sum(x)/length(x)})
```

## Data Visualization
### Distribution of Gender
```{r}
ggplot(ratings, aes(x = factor(gender), fill = factor(gender))) +
  geom_bar( show.legend=FALSE) +
  xlab("Gender") +
  ylab("Total Count") +
  ggtitle("Distribution of Gender")
```

Ratings by Gender
```{r}
# Ratings by Gender
ggplot(subset(ratings, !is.na(gender)), aes(x = gender, fill = as.factor(rating))) +
  geom_bar() +
  ggtitle("Rating Count by Gender") +
  xlab("Gender") +
  ylab("Total Count") +
  labs(fill = "Survived")
```

Average Rating by Gender
```{r}
male_mean <- ratings %>% filter(gender=='M') %>% pull(rating) %>% mean
female_mean <- ratings %>% filter(gender=='F') %>% pull(rating) %>% mean
mean_gender <- c(male_mean, female_mean)
gender <- c("male","female")
mean_gender_df <- data.frame(gender, mean_gender)
ggplot(mean_gender_df, aes(x=gender, y=mean_gender)) + 
  geom_bar(stat="identity") +
  ggtitle("Average Rating by Gender") +
  xlab("Gender") +
  ylab("Average Rating")
```

Average Rating by Genre
```{r}
# convert genres to factor
genres <- ratings[,7:24]
for(i in 1:ncol(genres)) {
  genres[,i] <- as.factor(genres[,i])
}
ratings[,7:24] <- genres
genres_rating <- cbind(genres, ratings[,3])
colnames(genres_rating)[19] <- "rating"

# show average rating by genre
mean <- rep(0,ncol(genres_rating)-1)
for(i in 1:(ncol(genres_rating)-1)) {
  mean[i] <- genres_rating %>% filter(genres_rating[[i]] == 1) %>% pull(rating) %>% mean
}
genres <- names(genres)
df <- data.frame(genres, mean)
ggplot(df, aes(x=genres, y=mean)) + 
  geom_bar(stat="identity") + 
  coord_flip() + 
  xlab("Genre") +
  ylab("Average Rating") +
  ggtitle("Average Rating by Genre")
```


Rating by Age Group
```{r}
ggplot(ratings, aes(x = age_group, fill = as.factor(rating))) +
  geom_bar() +
  ggtitle("Rating Count by Age") +
  xlab("Age") +
  ylab("Total Count") +
  labs(fill = "Rating")
```

## Split train and test ####################################################
```{r, eval=FALSE}
# Split ratings back into train and test
ratings_train <- merge(ratings_train, ratings, by=names(ratings_train))
ratings_test <- merge(ratings_test, ratings, by=names(ratings_test))
```

## Model Training ###########################################################
```{r, eval=FALSE}
lm(rating ~., data=ratings[,-c(4,5)])
base_model <- lm(rating ~., data=ratings[,-c(1,2,4,5,29:32)])
summary(base_model)
model1 <- lm(rating ~., data=ratings[,-c(4,5,27:32)]) # .34
model2 <- lm(rating ~., data=ratings[,-c(1,4,5,27:32)]) # .2112
model3 <- lm(rating ~., data=ratings[,-c(2,4,5,27:32)]) # .28 
```

